---
title: "ILHSXC"
author: "Maureen Renaud"
date: '2022-07-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Illinois High School Cross-Country

Review of Illinois HS XC and Track

### Importing the data

The data for this project was obtained from www.ihsa.org.
The data for each race was copied onto a page in an Excel workbook.  This workbook, titled "ILHSXC" contains all of the results for each race (broken down by year, gender, and division) from 1996 to present day.

Using the packages "readxl" and "tidyverse", I was able to import all pages from this workbook into one data.

```{r}
library("readxl")
library("tidyverse")

#Reading in all sheets from the Excel Spreadsheet

path <- "ILHSXC.xlsx"
all_data <- path %>% excel_sheets() %>%
            set_names() %>%
            map_dfr(read_excel, path = path)
```

An initial look at the summary shows us that we need to clean up some of our variables so that they can be appropriately analyzed.

```{r}
summary(all_data)
```


The first step is to transform some of the character variables to factors.  This will enable us to use these variable as categorical variables in our statistical models.
First, I will convert 'Grade' to a factor containing Freshman, Sophomore, Junior, and Senior. Prior to uploading this data set, there was a fairly large number of results containing no grade.  I was able to fill in many of these by looking at previous years results, results from other races, and googling the name.  Unfortunately, there were still five entries that I could not find information on.  
When it comes to statistical analysis based on grade, I will have to eliminate these rows.  Fortunately, these five entries are equal to 0.02% of the total data set of 26654 entries and thus will not have a significant impact on the results.

```{r}
# Convert Grade To A Factor
#Analyze NAs for Grade 
all_data$Grade <- factor(all_data$Grade, levels = c("Fr", "So", "Jr", "Sr"))
all_data[which(is.na(all_data[ , 8])),]
```

Next, I will convert Gender, Division, and Year to factors.  Fortunately, I do not have any missing data for these factors.

```{r}
#Convert Gender to A Factor
all_data$Gender <- factor(all_data$Gender, levels = c("Female", "Male"))

#Convert Division to A Factor
all_data$Division <- factor(all_data$Division, levels = c("A", "AA", "1A", "2A", "3A"))

#Convert Year to a factor
all_data$Year <- as.factor(all_data$Year)
```


```{r}

XCData[,3] <- sapply(strsplit(XCData[,3],":"),
        function(x) {
             x <- as.numeric(x)
             x[1]+x[2]/60
         }
  )
```

## EDA



```{r echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
